---
title: "Get BLUPS combining all trial data"
author: "LucianoRogerio"
date: "2022-08-25"
output: workflowr::wflow_html
editor_options:
  chunk_output_type: console
---

## Previous step
Prepare training dataset: Download data from DB, “Clean” and format DB data.


## Get multi-trial BLUPs from raw data (two-stage)

Two-stage procedure:

1. Fit mixed-model to multi-trial dataset and extract BLUPs, de-regressed BLUPs and weights. Include two rounds of outlier removal.
2. Genomic prediction with drg-BLUPs from multi-trial analysis as input.

**Work below represents Stage 1 of the Two-stage procedure.**

To fit the mixed-model I used last year, I am again resorting to asreml. I fit random effects for rep and block only where complete and incomplete blocks, respectively are indicated in the trial design variables. `sommer` should be able to fit the same model via the at() function, but I am having trouble with it’s memory intnsity and sommer is much slower even without a dense covariance (i.e. a kinship), compared to `lme4::lmer()` or `asreml::asreml()`.

To use `asreml` I require to access the license available only on robbins servers, I am actually using one called cbsumezey03.biohpc.cornell.edu.

This is the only step like this in the pipeline.

cbsumezey03 is using a SLURM job scheduler now. According to instructions, start an interactive bash shell with requested amount of recources, as follows:

```{bash Especify the hardware to be used in analysis, eval = F}
screen -R Luciano;
cd /workdir/lbraatz/IITA_2021GS/;
salloc -n20  --mem=100G;
# salloc: Granted job allocation 833
# salloc: Waiting for resource configuration
# salloc: Nodes cbsurobbins are ready for job
R;

```


## Set-up training datasets

```{r Prepare parameters, eval = F}
library(tidyverse); library(magrittr);
library(genomicMateSelectR)
dbdata<-readRDS(here::here("output","IITA_ExptDesignsDetected_2022Aug24.rds"))
traits<-c("MCMDS","DM","PLTHT","BRNHT1","BRLVLS","HI",
          "logDYLD", # <-- logDYLD now included. 
          "logFYLD","logTOPYLD","logRTNO","TCHART","LCHROMO","ACHROMO","BCHROMO")

# **Nest by trait.** Need to restructure the data from per-trial by regrouping by trait. 
dbdata<-nestDesignsDetectedByTraits(dbdata,traits)

dbdata %>% mutate(N_blups=map_dbl(MultiTrialTraitData,nrow)) %>% rmarkdown::paged_table()

dbdata %<>%
  mutate(fixedFormula=ifelse(Trait %in% c("logDYLD","logFYLD","logRTNO","logTOPYLD"),
                             "Value ~ yearInLoc + PropNOHAV","Value ~ yearInLoc"),
         randFormula=paste0("~idv(GID) + idv(trialInLocYr) + at(CompleteBlocks,'Yes'):repInTrial ",
                            "+ at(IncompleteBlocks,'Yes'):blockInRep"))
dbdata %>%
  mutate(Nobs=map_dbl(MultiTrialTraitData,nrow)) %>%
  select(Trait,Nobs,fixedFormula,randFormula) %>%
  rmarkdown::paged_table()
```

## Function to run asreml
Includes rounds of outlier removal and re-fitting.
Updated for new asreml arguments and output files.

```{r Mixed Model Function, eval = F}
fitASfunc<-function(fixedFormula,randFormula,MultiTrialTraitData,...){
  # test arguments for function
  # ----------------------
  # MultiTrialTraitData<-dbdata$MultiTrialTraitData[[7]]
  # #Trait<-dbdata$Trait[[7]]
  # fixedFormula<-dbdata$fixedFormula[[7]]
  # randFormula<-dbdata$randFormula[[7]]
  #test<-fitASfunc(fixedFormula,randFormula,MultiTrialTraitData)
  # ----------------------
  MultiTrialTraitData %<>%
    mutate(across(c(GID,yearInLoc,
                    CompleteBlocks,
                    IncompleteBlocks,
                    trialInLocYr,
                    repInTrial,
                    blockInRep),as.factor)) %>% 
    droplevels
  
  require(asreml); 
  fixedFormula<-as.formula(fixedFormula)
  randFormula<-as.formula(randFormula)
  # fit asreml 
  out<-asreml(fixed = fixedFormula,
              random = randFormula,
              data = MultiTrialTraitData, 
              maxiter = 40, workspace=1000e6, 
              na.action = na.method(x = "omit"))
  #### extract residuals - Round 1
  
  outliers1<-which(abs(scale(out$residuals))>3.3)
  
  if(length(outliers1)>0){
    
    x<-MultiTrialTraitData[-outliers1,]
    # re-fit
    out<-asreml(fixed = fixedFormula,
                random = randFormula,
                data = x, 
                maxiter = 40, workspace=1000e6, 
                na.action = na.method(x = "omit"))
    #### extract residuals - Round 2
    outliers2<-which(abs(scale(out$residuals))>3.3)
    if(length(outliers2)>0){
      #### remove outliers
      x<-x[-outliers2,]
      # final re-fit
      out<-asreml(fixed = fixedFormula,
                  random = randFormula,
                  data = x, maxiter = 40,workspace=1000e6, 
                  na.action = na.method(x = "omit"))
    }
  }
  if(length(outliers1)==0){ outliers1<-NULL }
  if(length(outliers2)==0){ outliers2<-NULL }
  
  ll<-summary(out,all=T)$loglik
  varcomp<-summary(out,all=T)$varcomp
  Vg<-varcomp["GID!GID","component"]
  Ve<-varcomp["units!R","component"]
  H2=Vg/(Vg+Ve)
  blups<-summary(out,coef=T)$coef.random %>%
    as.data.frame %>%
    rownames_to_column(var = "GID") %>%
    dplyr::select(GID,solution,std.error) %>%
    filter(grepl("GID",GID)) %>%
    rename(BLUP=solution) %>%
    mutate(GID=gsub("GID_","",GID),
           PEV=std.error^2, # asreml specific
           REL=1-(PEV/Vg), # Reliability
           drgBLUP=BLUP/REL, # deregressed BLUP
           WT=(1-H2)/((0.1 + (1-REL)/REL)*H2)) # weight for use in Stage 2
  out<-tibble(loglik=ll,Vg,Ve,H2,
              blups=list(blups),
              varcomp=list(varcomp),
              outliers1=list(outliers1),
              outliers2=list(outliers2))
  gc()
  return(out) }
```

## Run asreml

Ran in small chunks. Still learning SLURM scheduler used on server.

```{r Asreml Model, eval = F}
library(asreml)

require(furrr); plan(multisession, workers = 4)
options(future.globals.maxSize=+Inf); options(future.rng.onMisuse="ignore")
test<-dbdata %>%
  mutate(fitAS=future_pmap(.,fitASfunc))
saveRDS(test,file=here::here("output","test_2022Aug25.rds"))
plan(sequential)
rm(test); gc()

dbdata<-readRDS(here::here("output","test_2022Aug25.rds")) %>% 
  select(-fixedFormula,-randFormula,-MultiTrialTraitData)
dbdata %<>%
  unnest(fitAS)
```

## Output file

```{r Saving Results, eval = F}
saveRDS(dbdata,file=here::here("output","IITA_blupsForModelTraining_twostage_asreml_2022Aug25.rds"))
```

## [Home](index.html)

## [Last step]

1.2 [Phenotyping Data Curation](cleanTPdata.html)

## Results

See [Results](07-Results.html): Home for plots and summary tables.

## Next step

3. [Validate the pedigree obtained from cassavabase](03-validatePedigree.html): Before setting up a cross-validation scheme for predictions that depend on a correct pedigree, add a basic verification step to the pipeline. Not trying to fill unknown or otherwise correct the pedigree. Assess evidence that relationship is correct, remove if incorrect.

